---
title: "Profit from Policing : Allocation of Civil Asset Forfeiture Funds"
author: "Yuhang Li"
date: "7/10/2023"
output: github_document
---

```{r load packages}
#load required packages 
library(tidyverse)
library(knitr)
library(scales)
library(huxtable)
```

```{r load datasets}
#load cleaned data sets 
national_rev <- read.csv("forfeit_cleaned/national_rev_cleaned.csv")
national_exp <- read.csv("forfeit_cleaned/national_exp_cleaned.csv")
IL_cleaned <- read_csv("forfeit_cleaned/IL_cleaned.csv")
```

# Introduction 

Civil asset forfeiture is one of the most controversial policies in the United States. Civil asset forfeiture is a legal process that allows police officers to take and keep property that they suspect has been involved in criminal activity. Under this policy, police officers do not need to provide evidence that the original owner has committed a crime. Proponents of the policy state that civil asset forfeiture acts as a law enforcement tool to reduce material-related crime such as drug sales or vehicle crimes. However, the policy has also been criticized for its abusive use by police officers who seize cash and property from innocent civilians with the incentive to generate revenue. Therefore, it is important for policy makers to determine the pattern of civil asset forfeiture funds. What trends does revenue generated from civil asset forfeiture follow, and how are forfeiture funds spent? 

This data analysis report aims to examine patterns in expenditure and revenue of civil asset forfeiture funds over time. The report uses national and state level revenue and expense data to answer the research question. Results may lead to more insights on the financial incentives behind civil asset forfeiture laws and the impact of forfeiture revenue on budgets of law enforcement agencies. 

# Data

The [Institute for Justice](https://ij.org/report/policing-for-profit-3/policing-for-profit-data/) provides federal and state level data of civil asset forfeiture cases from 1986 - 2019. The data sets contain information about the type of property forfeited, the amount of revenue gained from the forfeiture, and the expenditure amount and allocation of forfeiture funds. Unfortunately, data is not available for all states and many states are not simultaneously present in both revenue and expense data. Thus, I chose to conduct the majority of my analysis at the national level by which more data is available. In addition, I will analyze fund patterns in Illinois, the state with the highest number of forfeiture cases. 

# Data Analysis 

## Part 1 : National Revenue and Expenditure of Civil Asset Forfeiture Funds (1986 - 2019)

Civil asset forfeiture existed for a long time in the United States but attracted increased attention in the 1980s as illegal drug activity increased in frequency. The right to seize materials from people with suspected involvement in criminal activity was aimed at reducing further illegal activity or trade. The graphs below show trends of the frequency of civil asset forfeiture cases in the United States from the period 1986 to 2019. 

```{r}
#plot number of civil asset forfeiture cases over years 
forfeit_nb_years <- national_rev %>% 
  ggplot(mapping = aes(x = year)) +
  geom_freqpoly(color = "#144a74") +
  labs(title = "Number of civil asset forfeiture cases over time",
       x = "Year",
       y = "Number of civil asset forfeiture cases") + 
  theme_classic()   
forfeit_nb_years 
```
National data shows that the number of civil asset forfeiture cases remained relatively stable until 1999. After that, from 2000 - 2019, the number of cases increased almost exponentially to over 100,000 cases. The number of cases peaked in 2018 to 125,000 cases. This could be attributed to many factors such as increased law enforcement activity, rapid policy changes related to civil asset forfeiture, financial motivations, or increased public attention on this topic in recent years. 


```{r average revenue per year, warning=FALSE}
#plot a bar graph of average revenue over years  
avg_rev_year <- national_rev %>% 
  group_by(year) %>% 
  summarize(avg_rev = mean(rev, na.rm = TRUE)) %>% 
  ggplot(mapping = aes(x = year, y = avg_rev)) +
  geom_bar(stat = "identity", fill = "#144a74") +
  labs(title = "Average revenue from civil asset forfeiture (1986 - 2019)",
       x = "Year",
       y = "Average revenue in U.S. dollars") +
  theme_classic()
avg_rev_year 
```
The graph above shows that the average revenue from civil asset forfeiture was extraordinarily high in 2014 and 2015 and averaged to over 200,000 dollars in each of the two years. This may be due to policy changes and regulations such as the Civil Asset Forfeiture Reform Act of 2014 that refines the scope and conditions of civil asset forfeiture. Law enforcement agencies could be more active in those years in targeting and seizing asset related to criminal activities, leading to a spike in revenues. If more time was given, I would conduct more detailed research on changes to civil asset forfeitures laws in 2014 and 2015 in different states.  

```{r proportion of total exp}
#write a function to find proportion of total expenditure for each expenditure type 
exp_prop <- function(dataset) {
  exp_prop_calc <- dataset %>% 
     #filter by positive expenditure values 
    filter(exp_type != "NA" & exp_amt > 0) %>%
    group_by(exp_type) %>% 
    summarize(total_exp = sum(exp_amt)) %>% 
     #calculate proportion of total expenditure 
    mutate(exp_pct = total_exp / sum(total_exp))
  return(exp_prop_calc)
}

#find proportion of national expenditure 
national_exp_type <- exp_prop(national_exp)

#plot proportion of total expenditure by type of expenditure
exp_type_bar <- national_exp_type %>% 
  ggplot(aes(x = reorder(exp_type, exp_pct), y = exp_pct, fill = exp_type)) +
  geom_bar(stat = "identity") +
  coord_flip() + 
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Proportion of forfeiture expenditure by expenditure type",
       x = "Proportion of total expenditure",
       y = "Expenditure Type",
       fill = "Expenditure Type") +
  theme_classic()
exp_type_bar
```
An analysis of expenditure data on the national level shows that the highest proportion of total expenditure is spent on capital expenditures (0.36) and the lowest proportion is spent on victim compensation (0.001). This suggests that almost no funds are allocated to services for victims and communities and that most of the funds gained from forfeiture are allocated to improving the facilities or equipment of law enforcement agencies. In the case that property is taken from innocent individuals, the revenue is not paid back to the community but instead directed to the finances of law enforcement agencies. 

The national level data set also provides information on whether the individuals whose property is confiscated end up being convicted or not. The following table shows the percentage of cases in which the owner involved was convicted of crime after confiscation of their property. 

```{r percent convicted}
#find proportion of cases with positive convicted outcome by property type 
percent_convicted <- national_rev %>%
  group_by(prop_type) %>% 
  summarize(percent_con = mean(conviction, na.rm = TRUE)) %>% 
  filter(prop_type != "NA") %>% 
  arrange(desc(percent_con))

#format as table 
percent_convic_tab <- percent_convicted %>% 
  #convert to percentage  
  mutate(percent_con = percent(percent_con, accuracy = 0.01)) %>% 
  kable(caption = "Percentage of civil asset forfeitures with convicted outcomes",
        col.names = c("Property type",
                      "Percent convicted after seizure")
  )    
percent_convic_tab
```

As shown by the table, only a minority of civil asset forfeitures involved cases with owners eventually convicted of crime. The proportion of positive conviction outcomes is around three times higher for real property seizures than for currency and and vehicle seizures. This indicates that most suspected seizures, especially currency and vehicles, are actually from innocent individuals. These individuals often do not have the resources or means to file a case and regain their lost property. 

Since different states have different laws regarding civil asset forfeiture, it may be useful to examine revenue and expense patterns at the state level. The following graph shows the top 10 states that had the most occurrences of civil asset forfeiture and the 10 states that had the least occurrences. 

```{r most and least cases}
#find the top 10 and bottom 10 states that had the most and least 
#civil asset forfeitures         
#find number of cases per state 
cases_per_state <- national_rev %>% 
  group_by(state) %>% 
  summarize(count = n()) %>% 
  filter(count > 60) %>% 
  arrange(desc(count)) 

#find top 10 states with most cases
top_10 <- cases_per_state %>% 
  head(10)

#find bottom 10 states with least cases
bottom_10 <- cases_per_state %>% 
  tail(10)

#combine the two groups
combined_states <- rbind(top_10, bottom_10)

#differentiate the two groups
combined_states$group <- if_else(combined_states$state %in%
                                   top_10$state,
                                 "Top 10",
                                 "Bottom 10")

#plot bargraph 
forfeit_rank_states <- combined_states %>% 
  ggplot(mapping = aes(x = reorder(state, count), y = count, fill = group)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("Top 10" = "#599e94",
                               "Bottom 10" = "#d2980d")) +
  labs(title = "Civil asset forfeiture cases 1986 - 2019",
       subtitle = "States that had most and least number of cases",
       x = "Number of civil asset forfeiture cases",
       y = "State",
       fill = "Group") + 
  theme_classic()

#print bargraph 
forfeit_rank_states       
```         

The graph shows states that has the most and least number of civil asset forfeiture cases recorded in the data during the period of 1986 - 2019. Interestingly, there is a large difference in the top ten and bottom ten states. The ten states that had the least number of civil asset forfeitures only had several hundred cases, while the ten states that had the most civil asset forfeitures generally had 25,000 to 100,000 cases. The state with the highest number of occurrences is Illinois with 98153 cases and the state with the lowest number of occurrences is West Virginia with only 114 cases. This suggests the existence of a wide disparity in civil asset forfeiture occurrences in the United States. Civil asset forfeiture is much more popular in some states than in others, which is potentially due to local law enforcement regulations or differences in state legislature. 

The provided state level data sets do not contain the expense and revenue data simultaneously for some states. Since Illinois ranked the highest in the number of civil asset forfeitures and has complete state level data, I decided to closely examine civil asset forfeiture trends at the state level for Illinois.

## Part 2 : Case Study of Illinois

Currently, according to [Illinois Policy Reports](https://www.illinoispolicy.org/reports/asset-forfeiture-in-illinois/), state law in Illinois authorizes the forfeiture of property, vehicles, vessels and aircraft, currency, and drugs on the basis of the property allegedly having been used in the commission of any crimes. A large percentage of revenue profits state agencies with the rest shared by the federal government. Reform policies are in progress to provided more transparency about how forfeiture funds are acquired and used. 

```{r}

selected_cols <- c("revenue", "exp_amt", "agency_share") 

#for loop to calculate summary statistics for revenue, expenditure amount, 
summary_stats <- list()
for (col in selected_cols) {
  column_summary <- summary(IL_cleaned[[col]], na.rm = TRUE)
  summary_stats[[col]] <- column_summary
}

summary_table <- hux(Variable = c("revenue", "expenditure", "agency share"),
                     Min = c(0, 0, 0),
                     Q1 = c(278, 0, 64.2),
                     Median = c(714, 0, 175),
                     Q3 = c(2094, 0, 538.4),
                     Max = c(17396274, 3189760, 86450)) %>% 
  set_all_padding(4) %>%
  set_outer_padding(0) %>%
  set_number_format(0) %>%
  set_bold(row = 1, col = everywhere) %>%
  set_bottom_border(row = 1, col = everywhere) %>%
  set_width(0.4) %>%
  set_caption("Summary statistics for revenue, expenditure, and agency share")

summary_table
```
The above table shows the summary statistics for the revenue gained from forfeiture, expenditure amount of forfeiture funds, and the amount of revenue the agency gains in Illinois. The median and the maximum value differ by an extreme amount, which suggests that the distribution of the three variables is highly skewed and has a lot of outliers. Assuming correct data entry, this could indicate that revenue and expenditure funds in Illinois vary greatly between different agencies. For instance, Chicago Police Department has a relatively high amount of revenue compared to other agencies in Illinois. Size of agency, number of employees, and surveillance area can all influence the amount of funds gained and spent. 


```{r}
IL_avg <- IL_cleaned %>% 
  #remove outlier 
  filter(year != "2009") %>% 
  group_by(year) %>% 
  summarize(avg_rev = mean(revenue, na.rm = TRUE),
            avg_exp = mean(exp_amt, na.rm = TRUE))

IL_avg_rev <- IL_avg %>% 
  ggplot(mapping = aes(x = year, y = avg_rev)) +
  geom_freqpoly(stat = "identity", color = "#144a74") +
  labs(title = "Average revenue from IL civil asset forfeiture (2000 - 2019)",
       x = "Year",
       y = "Average revenue in U.S. dollars") +
  theme_classic()

IL_avg_rev
```

The graph above analyzes the trend of average revenue over the years provided in the dataset, from 2000 - 2019. Unlike the national average revenue trend in part 1, the average revenue from civil asset forfeiture in Illinois remained relatively high until 2007 then decreased abruptly and remained at low values from 2010 - 2019. One explanation of this drop in average revenue may be stricter regulations regarding forfeitures. In recent years, seizures and forfeitures must be publicly reported and a substantial proof of crime upon seizure is emphasized by the government. 

Since distribution of variables are highly skewed, analysis of expenditure is better done by calculating the proportion of total expenditure of each type of expense and visualizing the results in a pie chart. 
```{r}

IL_exp_type <- exp_prop(IL_cleaned)

IL_exp_type_pie <- IL_exp_type %>% 
  ggplot(aes(x = "", y = exp_pct, fill = exp_type)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y") +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  labs(title = "Proportion of total expenditure by type",
       x = NULL,
       y = NULL,
       fill = "Type of Expenditure")
IL_exp_type_pie
```
The above pie chart shows the proportion of total expenditure of different types of expenditure areas. Similar to results from the national data, it is clear that the majority of forfeiture funds in Illinois is spent on capital expenditures. A small proportion is spent on community programs. This suggests that civil asset forfeiture is a crucial component of the finances of local agencies and largely contributes to the purchase of equipment and other capital projects. 

#Conclusion 

Every year, law enforcement agencies gain large amounts of revenue from cash, vehicles, and property confiscations from civilians. Data analysis results at the national level and the state level for Illinois both show that civil asset forfeiture funds are important components for financing capital expenditures of these agencies. However, at the national level, the majority of civil asset forfeiture cases did not involve individuals who are guilty of crime. The heavy allocation of forfeiture funds to capital expenditures may introduce incentives for law enforcement agencies to engage in "policing for profit", which harms innocent property owners. 

# Caveats and Further Steps 

The analysis of this report may be subject to certain limitations. First, the provided national and state level data sets contained many "NA" values for each column. This may be caused by errors in data collecting and may affect the results of the data analysis to a certain degree.  

Secondly, it is likely that the type of property being confiscated impacts the need for civil asset forfeiture laws. For instance, cases involving drugs and other illegal substances may require civil asset forfeiture to a larger degree than cases involving currency or vehicles. Civil asset forfeiture of illegal substances may be particularly effective in preventing further dissemination and harm. Even if the owner of illegal drugs is not convicted of crime, confiscation of the substance likely reduces its presence in the country. The provided data set does not specify what the "other" category of property type contains, so cases involving drug activity cannot be analyzed. Given more time, I would research more on the relationship between civil asset forfeiture and different types of property forfeited. 

#References 
[Institute for Justice](https://ij.org/report/policing-for-profit-3/policing-for-profit-data/)

[Illinois Policy Reports](https://www.illinoispolicy.org/reports/asset-forfeiture-in-illinois/)

[Data is Plural](https://www.data-is-plural.com/) 


