---
title: "Data Analysis for Profit from Policing : Allocation of Civil Asset Forfeiture Funds"
author: "Yuhang Li"
date: "7/10/2023"
output: github_document 
---

```{r load packages, warning=FALSE}
#load required packages 
library(tidyverse)
library(knitr)
library(scales)
library(huxtable)
```

```{r load datasets}
#load cleaned data sets 
national_rev <- read.csv("forfeit_cleaned/national_rev_cleaned.csv")
national_exp <- read.csv("forfeit_cleaned/national_exp_cleaned.csv")
IL_cleaned <- read_csv("forfeit_cleaned/IL_cleaned.csv") 
```

```{r check datasets}
#check data sets 
head(national_rev)
head(national_exp)
head(IL_cleaned)
```
#Part 1: National level data 
## Plot number of civil asset forfeiture cases per year 
```{r cases per year, warning=FALSE}
#plot number of civil asset forfeiture cases over years using line graph 
forfeit_nb_years <- national_rev %>% 
  ggplot(mapping = aes(x = year)) +
  geom_freqpoly(color = "#144a74") +
  labs(title = "Number of civil asset forfeiture cases over time",
       x = "Year",
       y = "Number of civil asset forfeiture cases") + 
  theme_classic()   
forfeit_nb_years 
```

## Analyze the trend of average revenue from civil asset forfeiture fund from 1986 to 2019
```{r average revenue per year, warning=FALSE}
#plot a bar graph of average revenue over years 
avg_rev_year <- national_rev %>% 
  group_by(year) %>% 
  summarize(avg_rev = mean(rev, na.rm = TRUE)) %>% 
  ggplot(mapping = aes(x = year, y = avg_rev)) +
  geom_bar(stat = "identity", fill = "#144a74") +
  labs(title = "Average revenue from civil asset forfeiture (1986 - 2019)",
       x = "Year",
       y = "Average revenue in U.S. dollars") +
  theme_classic()
avg_rev_year 
```

## Analyze the proportion of total expenditure for each expenditure type 
```{r proportion of total exp}
#write a function to find proportion of total expenditure for each expenditure type 
exp_prop <- function(dataset) {
  exp_prop_calc <- dataset %>% 
    #filter by positive expenditure values 
    filter(exp_type != "NA" & exp_amt > 0) %>%
    group_by(exp_type) %>% 
    summarize(total_exp = sum(exp_amt)) %>% 
    #calculate proportion of total expenditure 
    mutate(exp_pct = total_exp / sum(total_exp))
  return(exp_prop_calc)
}

#find proportion of national expenditure 
national_exp_type <- exp_prop(national_exp)

#plot proportion of total expenditure by type of expenditure
exp_type_bar <- national_exp_type %>% 
  ggplot(aes(x = reorder(exp_type, exp_pct), y = exp_pct, fill = exp_type)) +
  geom_bar(stat = "identity") +
  coord_flip() + 
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Proportion of forfeiture expenditure by expenditure type",
       x = "Proportion of total expenditure",
       y = "Expenditure Type",
       fill = "Expenditure Type") +
  theme_classic()
exp_type_bar
```

## Find the percentage of civil asset forfeitures involving individuals convicted with crime, grouped by property type 
```{r percent convicted}
#find proportion of cases with positive convicted outcome by property type 
percent_convicted <- national_rev %>%
  group_by(prop_type) %>% 
  summarize(percent_con = mean(conviction, na.rm = TRUE)) %>% 
  filter(prop_type != "NA") %>% 
  arrange(desc(percent_con))

#format as table 
percent_convicted %>% 
  #convert to percentage 
  mutate(percent_con = percent(percent_con, accuracy = 0.01)) %>% 
  kable(caption = "Percentage of civil asset forfeitures with convicted outcomes",
        col.names = c("Property type",
                      "Percent convicted after seizure")
  )      
```
##Find the top 10 and bottom 10 states that had the most and least civil asset forfeitures cases 
```{r most and least cases}
#find number of cases per state 
cases_per_state <- national_rev %>% 
  group_by(state) %>% 
  summarize(count = n()) %>% 
  filter(count > 60) %>% 
  arrange(desc(count)) 

#find top 10 states with most cases
top_10 <- cases_per_state %>% 
  head(10)

#find bottom 10 states with least cases
bottom_10 <- cases_per_state %>% 
  tail(10)

#combine the two groups
combined_states <- rbind(top_10, bottom_10)

#differentiate the two groups
combined_states$group <- if_else(combined_states$state %in%
                                   top_10$state,
                                 "Top 10",
                                 "Bottom 10")

#plot bargraph 
forfeit_rank_states <- combined_states %>% 
  ggplot(mapping = aes(x = reorder(state, count), y = count, fill = group)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("Top 10" = "#599e94",
                               "Bottom 10" = "#d2980d")) +
  labs(title = "Civil asset forfeiture cases 1986 - 2019",
       subtitle = "States that had most and least number of cases",
       x = "Number of civil asset forfeiture cases",
       y = "State",
       fill = "Group") + 
  theme_classic()

#print bargraph 
forfeit_rank_states       
```         

#Part 2 : Illinois Data  
## 
```{r}
IL_rev_property <- IL_cleaned %>% 
  filter(prop_type != "NA" & revenue < 5000) %>% 
  ggplot(mapping = aes(x = prop_type, y = revenue)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Revenue by type of property",
       x = "revenue",
       y = "property type") +
  theme_classic()
IL_rev_property   
```
```{r}

selected_cols <- c("revenue", "exp_amt", "agency_share") 

#for loop to calculate summary statistics for revenue, expenditure amount, 
summary_stats <- list()
for (col in selected_cols) {
  column_summary <- summary(IL_cleaned[[col]], na.rm = TRUE)
  summary_stats[[col]] <- column_summary    
}

summary_table <- hux(Variable = c("revenue", "expenditure", "agency share"),
                     Min = c(0, 0, 0),
                     Q1 = c(278, 0, 64.2),
                     Median = c(714, 0, 175),
                     Q3 = c(2094, 0, 538.4),
                     Max = c(17396274, 3189760, 86450)) %>% 
  set_all_padding(4) %>%
  set_outer_padding(0) %>%
  set_number_format(0) %>%
  set_bold(row = 1, col = everywhere) %>%
  set_bottom_border(row = 1, col = everywhere) %>%
  set_width(0.4) %>%
  set_caption("Summary statistics for revenue, expenditure, and agency share")

summary_table
```

```{r}
IL_avg <- IL_cleaned %>% 
  #remove outlier 
  filter(year != "2009") %>% 
  group_by(year) %>% 
  summarize(avg_rev = mean(revenue, na.rm = TRUE),
            avg_exp = mean(exp_amt, na.rm = TRUE))

IL_avg_rev <- IL_avg %>% 
  ggplot(mapping = aes(x = year, y = avg_rev)) +
  geom_freqpoly(stat = "identity", color = "#144a74") +
  labs(title = "Average revenue from IL civil asset forfeiture (2000 - 2019)",
       x = "Year",
       y = "Average revenue in U.S. dollars") +
  theme_classic()

IL_avg_rev

```



```{r}
IL_exp_type <- IL_cleaned %>% 
  group_by(exp_type) %>% 
  summarize(total_exp = sum(exp_amt)) %>% 
  filter(exp_type != "NA") %>% 
  mutate(exp_proportion = total_exp / sum(total_exp))
IL_exp_type 

IL_exp_type_pie <- IL_exp_type %>% 
  ggplot(aes(x = "", y = exp_proportion, fill = exp_type)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y") +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  labs(title = "Proportion of total expenditure by type",
       x = NULL,
       y = NULL,
       fill = "Type of Expenditure")
IL_exp_type_pie
```



